A self-consistency check for unitary propagation of Hawking quanta

### 2016-11-01 Draft_BH_shells.tex
Changes from today

\title{A self-consistency check for unitary Hawking radiation}

@@

\author{I-Sheng Yang}

@@

\begin{itemize}
\item {\bf Classical coherence by tracking geodesics.} The classical mode of a massless field basically has its peaks and nodes following null geodesics.
Thus, instead of actually solving the mode function, we can perform a much simpler calculation to track geodesics.
We start with two out-going null geodesics with the separation of one period of such mode near the horizon, and then we calculate the change in their separations at spatial infinity.
The change of distance (arrival time difference) between these two geodesics qualitatively represent the change a mode function will have.
As such change gets large, we should lose faith on the apparent unitarity in the QFT calculation.
\item {\bf The background and the uncertainty.} The classical background in this case is the Schwarzschild geometry, which can be written as the metric $g_{\mu\nu}$. 
A classical uncertainty can be expressed as deviation from such metric, $\Delta g_{\mu\nu}$. 
There are two major challenges here.
\begin{enumerate}
\item There are many more variables than the double-slit experiment. 
We consider only one particular form of $\Delta g_{\mu\nu}$ in this paper---back-reactions from the presence of extra matter of zero total energy. 
It is simple to calculate and reflects the physcial intuition of vacuum pair fluctuation.
\item The value of $\Delta g_{\mu\nu}$ is gauge dependent. 
By relating it to the presence of extra matter, we can express it as a gauge-invariant local quantity, which is the natural thing to hold fixed in our analysis.
\end{enumerate}
Note that if we were reporting a positive result, that such uncertainty upholds unitarity, then one should question the validity of learning a general lesson from a special example. However, we will show that this particular form of $\Delta g_{\mu\nu}$ is already sufficient to question unitarity, and we did not choose this form to specifically do that. Thus our limited analysis is sufficient to raise a reasonable doubt to the classical-background approximation.
\item {\bf Renormalization.} It is well-known that na\"ive applications of QFT often suffer from UV/IR divergences.
In our case, we will see that even in Minkowski space, geometric uncertainty already leads to an in-principle-finite change between the two null geodesics.
We will simply ``renormalize'' that away by saying that if the black hole case leads to a similar change, it should not be taken as a serious problem.
Our surprising result is that in the black hole case, we get an infinitely larger effects than in Minkowski space, which is difficult to blame on the usual divergence of QFT.
\end{itemize}

More concretely, our setup is shown in Fig.\ref{fig-setup}. 
We start by choosing two points at some $r_0\gg M$ with a $\delta t_0\sim M$ coordinate time separation between them.
This represents one wavelength of the mode function of an asymptotic Hawking quantum.
We then back-track two null geodesics from these two points back to two points near the horizon, at $r_A$ with $(r_A-2M)\ll 2M$.
Now their separation represents one wavelength of a near-horizon mode, which would have propagated to the asymptotic mode if there were no geometric uncertainty. 
We then introduce a geometric uncertainty as two shells of zero total ADM mass. 
Therefore, both the near horizon and the asymptotic regions are not affected by such uncertainty.
Only the process of propagation is affected.
With the same two starting points near the horizon. we calculate the separation $\bar{\delta t_0}$ of the two null geodesics when they arrive at $r_0$, and compare it with $\delta t_0$.

\begin{figure}[tb]
\begin{center}
\includegraphics[scale = 0.6]{Propertime.pdf}
\caption{The two parts of this diagrams describe two different scenarios. The trajectories of the photon are the lines in orange. The trajectory of the observer and the position from which the photons are emitted are represented by solid blue lines. The bottom part is a general case of two Hawking photons coming from a distance $r_A$ from the black hole of mass $M$ and arriving at an observer who is at a distance of $r_0$. The top part shows two Hawking photons coming from the same distance $r_A$ from a black hole of mass $M$. Instead of the photons freely propagating through to an observer at $r_0$, they have to cross two shells, represented by blue dotted line, of equal and opposite mass $\Delta M, -\Delta M$ at $r_{1}, r_{2}$ respectively. The regions $I$ represent a metric with mass $M$ and region $II$ represents a metric with mass $M-\Delta M$.}
\label{fig-setup}
\end{center}
\end{figure}

@@

\begin{equation}
	\frac{\delta t_0 - \bar{ \delta t_0}}{\delta t_0} = \frac{\Delta M \Delta r}{(r_1-2M)^2}	\label{lim-result}
\end{equation}

This expression can be evaluated in three limits; 1) The Minkowski limit ($M=0$), 2) Far from horizon ($M <<r_1$), 3) Near horizon limit ($r_1<<M$). These limits have to be calculated using physical, local quantities, which are the local surface energy of the shell $\sigma$ and the proper distances. The calculations are performed in appendix \ref{finalEq} and the results are summarized below:

\begin{equation}
\frac{\delta t_0 - \bar{ \delta t_0}}{\delta t_0} = \begin{cases}
 \sigma \Delta l~	 &\ \ \ \ \ (M = 0)\\
- 4\pi \sigma \Delta l \left( 1 + \frac{2M}{l} \right) & \ \ \ \ \ (M<<r_1)	\\
- 64 \pi  \sigma \Delta l \frac{M^2}{l^2} & \ \ \ \ \ (M>>r_1).
\end{cases}	\label{fin-result}
\end{equation}

@@

Many existing results of QFT in curved/dynamical geometries are about particle numbers, such as the particle-production calculation following a Bogoliugov transformation. 
As an example, one can easily see that
\begin{equation}
\rho_{\rm pure} = \left(\sum_n a_n|n\rangle\right)
\left(\sum_n \langle n| a_n^*\right)
\end{equation}
and
\begin{equation}
\rho_{\rm mixed} = \sum_n |a_n|^2 |n\rangle\langle n|
\end{equation}
give the same particle number. 
Thus there is no general crisis if we admit that local QFT sometimes loses subsystem unitarity.

@@

\acknowledgments

We thank XXX for discussions. We are supported by XXX fundings.

@@

\section{Deriving Eq (\ref{fin-result})}\label{finalEq}
\label{finalEq}

@@

\begin{equation}
	\frac{\delta t_0 - \bar{t_0}}{\delta t_0} = \frac{\Delta M \Delta l}{l^2} = \sigma \Delta l.
\end{equation}

\subsection{Far from horizon limit ($M<<r_1$)}

In this limit Eq (\ref{l-r}) can be expanded to first order to give

@@

\begin{eqnarray}
	\frac{\delta t_0 - \delta \bar{t_0}}{\delta t_0} & = & 4 \pi \sigma \frac{r^2}{(r_1 - 2M)^2} \left( 1 - \frac{2M}{r_1} \right)^{\frac{1}{2}} \left( 1 - \frac{2M}{r_1} \right)^\frac{1}{2}  \nonumber	\\
	& = & 4 \pi \sigma \Delta l \left( 1  + \frac{2M}{r_1 - 2M} \right)^2 \left( 1 - \frac{2M}{r_1} \right)	\nonumber	\\
	& = & 4 \pi \sigma \Delta l \left( 1 + \frac{2M}{l} \right) + \mathcal{O}(l^{-2})
\end{eqnarray}

@@

in this limit the coordinate distance $r_1$ and proper distance $l$ are related by

@@

\begin{eqnarray}
	\frac{\delta t_0 - \delta \bar{t_0}}{\delta t_0} & = &  4 \pi \sigma \delta l \left( 1+ \frac{2M}{r_1 - 2M} \right)^2 \left( 1 - \left( 1 + \frac{r_1 - 2M}{2M} \right)^{-1} \right)	\nonumber	\\
	& =& 4 \pi \sigma \delta l \left( \frac{2M}{r_1 - 2M}\right)^2 \left( \frac{r_1 - 2M}{2M}  + \mathcal{O}(M^{-2}) \right)	\nonumber	\\
	& = & 64 \pi \sigma \delta l \frac{M^2}{l^2}
\end{eqnarray}

>>>>

\title{A self-consistency check for unitary propagation of Hawking quanta}

@@

\author{I-Sheng Yang}s

@@

\begin{itemize}
\item {\bf Classical coherence by tracking geodesics.} The classical mode of a massless field basically has its peaks and nodes following null geodesics.
Thus, instead of actually solving the mode function, we can perform a much simpler calculation to track geodesics.
We start with two out-going null geodesics with the separation of one period of such mode near the horizon, and then we calculate the change in their separations at spatial infinity.
If this change is comparable to the wavelength, then the mode function of this wavelength loses classical coherence.
\item {\bf The background and the uncertainty.} The classical background in this case is the Schwarzschild geometry, which can be written as the metric $g_{\mu\nu}$. 
A classical uncertainty can be expressed as deviation from such metric, $\Delta g_{\mu\nu}$. 
There are two major challenges here.
\begin{enumerate}
\item There are many more variables than the double-slit experiment. 
We consider only one particular form of $\Delta g_{\mu\nu}$ in this paper---back-reactions from the presence of extra matter of zero total energy. 
It is simple to calculate and reflects the physcial intuition of vacuum pair fluctuation.
\item The value of $\Delta g_{\mu\nu}$ is gauge dependent. 
By relating it to the presence of extra matter, we can express it as a gauge-invariant local quantity, which is the natural thing to hold fixed in our analysis.
\end{enumerate}
Note that if we were reporting a positive result, that such uncertainty upholds unitarity, then one should question the validaty of learning a general lesson from a special example. However, we will show that this particular form of $\Delta g_{\mu\nu}$ is already sufficient to question unitarity, and we did not choose this form to specifically do that. Thus our limited analysis is sufficient to raise a reasonable doubt to the classical-background approximation.
\item {\bf Renormalization.} It is well-known that na\"ive applications of QFT often suffer from UV/IR divergences.
In our case, we will see that even in Minkowski space, geometric uncertainty already leads to a finite change between the two null geodesics.
We will simply ``renormalize'' that away by saying that if the black hole case leads to a similar change, it should not be taken as a serious problem.
Our surprising result is that in the black hole case, we get an infinitely larger change, which is difficult to blame on renormalization.
\end{itemize}

More concretely, our setup is shown in Fig.\ref{fig-setup}. 
We start by choosing two points at some $r_0\gg M$ with a $\delta t_0\sim M$ coordinate time separation between them.
This represents one wavelength of the mode function of an asymptotic Hawking quantum.
We then back-track two null geodesics from these two points back to two points near the horizon, at $r_A$ with $(r_A-2M)\ll 2M$.
Now their separation represents one wavelength of a near-horizon mode, which would have propagated to the asymptotic mode if there were no geometric uncertainty. 
We then introduce introduce a geometric uncertainty as two shells of zero total ADM mass. 
Therefore, both the near horizon and the asymptotic regions are not affected by such uncertainty.
Only the process of propagation is affected.
With the same two starting points near the horizon. we calculate the separation $\bar{\delta t_0}$ of the two null geodesics when they arrive at $r_0$, and compare it with $\delta t_0$.

\begin{figure}[tb]
\begin{center}
\includegraphics[scale = 0.6]{Propertime.pdf}
\caption{The two parts of this diagrams describe two different scenarios. The trajectories of the photon are the lines in orange. The trajectory of the observer and the position from which the photons are emitted are represented by solid blue lines. The bottom part is a general case of two Hawking photons coming from a distance $r_A$ from the black hole of mass $M$ and arriving at an observer who is at a distance of $r_0$. The top part shows two Hawking photons coming from the same distance $r_A$ from a black hole of mass $M$. Instead of the photons freely propagating through to an observer at $r_0$, they have to cross two shells, represented by blue dotted line, of equal and opposite mass $\Delta M, -\Delta M$ at $r_{1}, r_{2}$ respectively. The regions $I$ represent a metric with mass $M$ and region $II$ represents a metric with mass $M+\Delta M$.}
\label{fig-setup}
\end{center}
\end{figure}

@@

\begin{equation}
	\frac{\delta t_0 - \bar{ \delta t_0}}{\delta t_0} = \frac{\Delta M \Delta r}{(r_1-2M)^2} = \frac{\Delta M \Delta r}{r_1^2} g_{rr}^{2}	\label{lim-result}
\end{equation}

In fact, this expression can be written in general (see Eq (\ref{Eq-mink}) in the appendix for derivation) in terms of physical, local quantities, which are the local surface energy of the shell $\sigma$ and the proper distances,

\begin{equation}
	\frac{\delta t_0 - \bar{ \delta t_0}}{\delta t_0} = 4 \pi \sigma \Delta l g_{rr}
\end{equation}

where $g_{rr}$ is the radial component of the Schwasrchild metric. This expression can be evaluated in three limits; 1) The Minkowski limit ($M=0$), 2) Far from horizon ($M <<r_1$), 3) Near horizon limit ($r_1<<M$). The calculations are performed in appendix \ref{finalEq} and the results are summarized below:

\begin{equation}
\frac{\delta t_0 - \bar{ \delta t_0}}{\delta t_0} = \begin{cases}
 4 \pi \sigma \Delta l~	 &\ \ \ \ \ (M = 0)\\
4\pi \sigma \Delta l \left( 1 + \frac{2M}{l} \right) & \ \ \ \ \ (M<<r_1)	\\
64 \pi  \sigma \Delta l \frac{M^2}{l^2} & \ \ \ \ \ (M>>r_1).
\end{cases}	\label{fin-result}
\end{equation}

@@

Many existing results of QFT in curved/dynamical geometries are about particle numbers, such as the particle-production calculation following a Bogoliubov transformation. 
As an example, one can easily see that
\begin{equation}
\rho_{\rm pure} = \left(\sum_n a_n|n\rangle\right)
\left(\sum_n \langle n| a_n^*\right)
\end{equation}
and
\begin{equation}
\rho_{\rm mixed} = \sum_n |a_n|^2 |n\rangle\langle n|
\end{equation}
give the same particle number. 
Thus there is no general crisis if we admit that local QFT sometimes loses subsystem unitarity.

@@

@@

%\section{Deriving Eq (\ref{fin-result})}\label{finalEq}
\section{Whatever section name it was}
\label{finalEq}

@@

\begin{equation}
	\frac{\delta t_0 - \bar{t_0}}{\delta t_0} = \frac{\Delta M \Delta l}{l^2} = 4 \pi \sigma \Delta l.	\label{Eq-mink}
\end{equation}

Very far away from the horizon, we expect a small deviation in Eq (\ref{Eq-mink}) from the Minkowski result as we know that the Schwarzschild geometry is asymptotically flat. By assuming $r_1>>M$ we  expanded Eq \ref{l-r} to first order to give

@@

\begin{eqnarray}
	\frac{\delta t_0 - \delta \bar{t_0}}{\delta t_0} & = & 4 \pi \sigma \frac{r_1^2}{(r_1 - 2M)^2} \left( 1 - \frac{2M}{r_1} \right)^{\frac{1}{2}} \left( 1 - \frac{2M}{r_1} \right)^\frac{1}{2}  \nonumber	\\
	& = & 4 \pi \sigma \Delta l \left( 1 - \frac{2M}{r_1} \right)^{-1} = 4 \pi \sigma \Delta l g_{rr}	\label{gen-result}
\end{eqnarray}

Note that this is actually a general result so far in terms of the coordinate distance $r_1$. We can expand this to leading order now by assuming $2M/r_1 <<1$,

\begin{equation}
	\frac{\delta t_0 - \delta \bar{t_0}}{\delta t_0} = 4 \pi \sigma \Delta l \left( 1 + \frac{2M}{r_1} \right) 
\end{equation}

@@

In this limit the coordinate distance $r_1$ and proper distance $l$ are related by

@@

\begin{eqnarray}
	\frac{\delta t_0 - \delta \bar{t_0}}{\delta t_0} & = &  4 \pi \sigma \delta l \left( 1+ \frac{2M}{r_1 - 2M} \right)^2 \left( 1 - \left( 1 + \frac{r_1 - 2M}{2M} \right)^{-1} \right)	\nonumber	\\
	& =& 4 \pi \sigma \delta l \left( \frac{2M}{r_1 - 2M}\right)^2 \left( \frac{r_1 - 2M}{2M}  + \mathcal{O}(M^{-2}) \right)	\nonumber	\\
	& = & 64 \pi \sigma \delta l \frac{M^2}{l^2} + \mathcal{O}(r^{-2})
\end{eqnarray}
